---
title: "Hydro Outages - Marketing Data"
author: "M.C."
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
---

# check and load packages
```{r, echo=FALSE}
packages <- c("tm","stringr", "wordcloud", "caret", "readr", "dplyr", "tidyr", "datasets", "ggplot2", "scales", "lubridate", "fpp2", "zoo")

package.check <- lapply(packages, FUN = function(x) {
  if (!require(x, character.only = TRUE)) {
    install.packages(x, dependencies = TRUE)
    library(x, character.only = TRUE)
  }
})

# remove 'packages', 'package.check' after loading the required packages.
rm(packages)
rm(package.check)
```

# set directory
setwd("E:/Documents/OneDrive - Syracuse University/SU/Courses/IST 707/Project/Hydro Outages")

# ----------------------------------------------------------------------------------

# #################
# read in and combine sales data sets of different time periods
```{r}
sales2005 <- read.csv(file="m_transactions_to2005.csv", header=TRUE, sep=",", na=c("","NA"))
sales2005 <- sales2005[ ,-8:-25] # remove extra columns
sales2005$Year <- as.character(sales2005$Year)

sales2016 <- read.csv(file="m_transactions_to2016.csv", header=TRUE, sep=",", na=c("","NA"))
sales2011 <- read.csv(file="m_transactions_to2011.csv", header=TRUE, sep=",", na=c("","NA"))
sales2018mar <- read.csv(file="m_transactions_march2018.csv", header=TRUE, sep=",", na=c("","NA"))
sales2018end <- read.csv(file="m_transactions_end2018.csv", header=TRUE, sep=",", na=c("","NA"))

# combine sales2018to & sales2018end
sales2018 <- rbind(sales2011, sales2016, sales2018mar, sales2018end)
summary(sales2005)

# convert date variable into real date format
sales2005$TIME <- as.POSIXct(sales2005$TIME, format = "%m/%d/%Y %H:%M")
sales2018$TIME <- as.POSIXct(sales2018$TIME, format = "%m/%d/%Y %H:%M")

# extract years from sales2018, to match columns of sales2005
sales2018$Year <- format(sales2018$TIME, format="%Y")

# combine sales2005 and sales2018
sales <- rbind(sales2005, sales2018)
```

```{r}
# check data sets
str(sales)
summary(sales) 

# check rows with NA, and remove
sales[which(is.na(sales$TIME)),]
sales <- na.omit(sales)

# str(sales2005)
# str(sales2018)
# str(sales2011)
# str(sales2018end)

# remove progress data sets no longer needed
rm(sales2005)
rm(sales2011)
rm(sales2016)
# rm(sales2018)
rm(sales2018mar)
rm(sales2018end)
```


# ----------------------------------------------------------------------------------
# copy to a working data "trans" and create a Date variable by days
```{r}
# reserve sales as original sales data set, use strans as working set
trans <- sales

str(trans)

# create a Date variable - "Y-m-d"
trans$Date <- as.Date(trans$TIME)



```


# calculate sub-total of all transactions
```{r}
summary(trans)

# create a variable of daily sub-total
trans$subTotal <- trans$MWH * trans$PRICE

```




# ########### hourly transactions and prices ###########
# calculate hourly average prices of sales and purchases
```{r}
avgPrice_hr <- data.frame(tapply(trans$PRICE, list(trans$TIME,trans$PROVISIONTYPE), mean))
names(avgPrice_hr) <- c("avgPurchase.hr", "avgSale.hr")
avgPrice_hr$Time <- rownames(avgPrice_hr)
avgPrice_hr <- avgPrice_hr[ ,c(3,1,2)]

# replace NAs with 0
avgPrice_hr[is.na(avgPrice_hr)] <- 0

# convert Time variable to POSIXct objects 
avgPrice_hr$Time <- as.POSIXct(avgPrice_hr$Time, format = "%Y-%m-%d %H:%M")

```

# calculate hourly net-total
```{r}
netTotal_hr <- data.frame(tapply(trans$subTotal, trans$TIME, sum))
# netTotal_hr[c(1:10,101:110,17158:17168), ]
avgPrice_hr$netTotal.hr <- netTotal_hr[,1]

```

# calculate hourly volatilities
```{r}

# 1st, combine hourly average prices of sales and purchases to trans
trans <- left_join(trans, avgPrice_hr, by=c("TIME"="Time"))


# 2nd, separate Purchase and Sale under PROVISIONTYPE
trans.purchase <- trans[trans$PROVISIONTYPE=="Purchase", ]
trans.sale <- trans[trans$PROVISIONTYPE=="Sale", ]


# 3nd, calculate volatility of hourly price for Purchase and Sale respectively
avgPrice_hr$netTotal <- netTotal_hr[,1]

# normalized deviation of prices vs average hourly price
trans.purchase$ppVolatility.hr <- round(((trans.purchase$PRICE - trans.purchase$avgPurchase.hr) / trans.purchase$avgPurchase.hr), digits=3)
trans.sale$psVolatility.hr <- round(((trans.sale$PRICE - trans.sale$avgSale.hr) / trans.sale$avgSale.hr), digits=3)

# volatility of hourly prices for Purchase & Sale
# purchase
volPurchase_hr<- data.frame(tapply(trans.purchase$ppVolatility.hr, trans.purchase$TIME, mean))
names(volPurchase_hr) <- "avg.ppVolatility.hr"
volPurchase_hr$Time <- rownames(volPurchase_hr)
volPurchase_hr$Time <- as.POSIXct(volPurchase_hr$Time, format = "%Y-%m-%d %H:%M")
volPurchase_hr <- volPurchase_hr[ ,c(2,1)]

# sale
volSale_hr<- data.frame(tapply(trans.sale$psVolatility.hr, trans.sale$TIME, mean))
names(volSale_hr) <- "avg.psVolatility.hr"
volSale_hr$Time <- rownames(volSale_hr)
volSale_hr$Time <- as.POSIXct(volSale_hr$Time, format = "%Y-%m-%d %H:%M")
volSale_hr <- volSale_hr[ ,c(2,1)]

str(volPurchase_hr)
str(volSale_hr)


# 4th, put average hourly volitilities of Purchase and Sale prices to avgPrice_hr
avgPrice_hr <- left_join(avgPrice_hr, volPurchase_hr, by=c("Time"="Time"))
avgPrice_hr <- left_join(avgPrice_hr, volSale_hr, by=c("Time"="Time"))

str(avgPrice_hr)

```


# calculate hourly subtotal of purchase and sales
```{r}
prchsTotal_hr <- data.frame(tapply(trans.purchase$subTotal, trans.purchase$TIME, sum))
names(prchsTotal_hr) <- "prchsTotal.hr"
prchsTotal_hr$Time <- rownames(prchsTotal_hr)
prchsTotal_hr$Time <- as.POSIXct(prchsTotal_hr$Time, format = "%Y-%m-%d %H:%M")

saleTotal_hr <- data.frame(tapply(trans.sale$subTotal, trans.sale$TIME, sum))
names(saleTotal_hr) <- "saleTotal.hr"
saleTotal_hr$Time <- rownames(saleTotal_hr)
saleTotal_hr$Time <- as.POSIXct(saleTotal_hr$Time, format = "%Y-%m-%d %H:%M")

avgPrice_hr <- left_join(avgPrice_hr, prchsTotal_hr, by=c("Time"="Time"))
avgPrice_hr <- left_join(avgPrice_hr, saleTotal_hr, by=c("Time"="Time"))

```


# calculate hourly price variance of purchase and sales
```{r}
trans.purchase$ppVariance <- (trans.purchase$ppVolatility.hr)^2
trans.sale$psVariance <- (trans.sale$psVolatility.hr)^2

# purchase
variancePurchase_hr <- data.frame(tapply(trans.purchase$ppVariance, trans.purchase$TIME, sum))
names(variancePurchase_hr) <- "sum.ppVariance.hr"
variancePurchase_hr$Time <- rownames(variancePurchase_hr)
variancePurchase_hr$Time <- as.POSIXct(variancePurchase_hr$Time, format = "%Y-%m-%d %H:%M")

# sale
varianceSale_hr <- data.frame(tapply(trans.sale$psVariance, trans.sale$TIME, sum))
names(varianceSale_hr) <- "sum.psVariance.hr"
varianceSale_hr$Time <- rownames(varianceSale_hr)
varianceSale_hr$Time <- as.POSIXct(varianceSale_hr$Time, format = "%Y-%m-%d %H:%M")


avgPrice_hr <- left_join(avgPrice_hr, variancePurchase_hr, by=c("Time"="Time"))
avgPrice_hr <- left_join(avgPrice_hr, varianceSale_hr, by=c("Time"="Time"))

str(avgPrice_hr)
```


# calculate moving averages in a 24 hours window (use k=25 for symmetry)
```{r include=FALSE}
avgPrice_hr$mavPurchase.hr <- rollmean(avgPrice_hr$avgPurchase.hr, k = 25, fill = NA)
avgPrice_hr$mavSale.hr <- rollmean(avgPrice_hr$avgSale.hr, k = 25, fill = NA)

avgPrice_hr$mavPrchsTotal.hr <- rollmean(avgPrice_hr$prchsTotal, k = 25, fill = NA)
avgPrice_hr$mavSaleTotal.hr <- rollmean(avgPrice_hr$saleTotal, k = 25, fill = NA)
avgPrice_hr$mavNetTotal.hr <- rollmean(avgPrice_hr$netTotal, k = 25, fill = NA)
```

# write out the avgPRice_hr to a .csv file
```{r}
avgPrice_hr[is.na(avgPrice_hr)] <- 0

write.csv(avgPrice_hr, file = "market_hourly_data.csv")

```




# ########### daily transactions and prices ###########
# calculate average daily prices
```{r}
# what is the price range?
(max(trans$PRICE) - min(trans$PRICE))

# what is the average of daily prices?
avgPrice_daily <- data.frame(tapply(trans$PRICE, trans$Date, mean))
avgPrice_daily$date <- as.Date(rownames(avgPrice_daily), "%Y-%m-%d")
names(avgPrice_daily) <- c("avg_price","date")


str(avgPrice_daily)
head(avgPrice_daily)

# plot a time serie of average daily prices
ggplot(data=avgPrice_daily, aes(x=date, y=avg_price, group=1)) + 
  geom_line(color="steelblue", size=1, alpha=0.5) + labs(x="Year", y="Average Price") +
  scale_x_date(name="Year", date_breaks="1 year", date_labels="%Y") + ggtitle("Average Daily Price")
  
```

# combine avgPrice_daily to trans
```{r}
trans <- left_join(trans, avgPrice_daily, by=c("Date"="date"))

```

# calculate volatility of daily price
```{r}
# normalized deviation of prices vs average daily price
trans$pVolatility <- round(((trans$PRICE - trans$avg_price) / trans$avg_price), digits=3)

# volatility coef. of daily prices
volPrice_daily <- data.frame(tapply(trans$pVolatility, trans$Date, mean))
names(volPrice_daily) <- "avg_volatile"

volPrice_daily2 <- data.frame(tapply(trans$pVolatility, trans$Date, sum))
names(volPrice_daily2) <- "sum_volatile"

volPrice_daily$sum_volatile <- volPrice_daily2$sum_volatile

volPrice_daily$avg_volatileX <- round(volPrice_daily$avg_volatile * 1000000, digits=3)
volPrice_daily$sum_volatileX <- round(volPrice_daily$sum_volatile * 1000, digits=3)
volPrice_daily$date <- as.Date(rownames(volPrice_daily), "%Y-%m-%d")

head(volPrice_daily)

str(volPrice_daily)

rm(volPrice_daily1)
rm(volPrice_daily2)

# plot a time serie of votalities of daily prices
ggplot(data=volPrice_daily, aes(x=date, y=avg_volatileX, group=1)) + 
  geom_line(color="tomato", size=1, alpha=0.5) + labs(x="Year", y="Price Volatility Index") +
  scale_x_date(name="Year", date_breaks="1 year", date_labels="%Y") + ggtitle("Daily Price Volatility")
  

```
  

