---
output:
  pdf_document: default
  html_document: default
---
# IST 707 - Final Project
# Nicole Gruet-Matthews  
# SVMs, DT, NB, and RandomForest Predictor
# Can we predict class of outage (good, bad, or none)?  
# Read in the preprocessed csv
```{r}
## Set the WD and read in the Train/Test data
setwd <- ("C:/Users/ngruet/Desktop/Personal/Grad School/IST 707/project")
df <- read.csv("preprocesseddata_12.03.2019.csv", header = TRUE)

## What does our data look like?
str(df)

## We don't need all these variables
## Change row 1 to row names and delete row 1 (dateTime)
row.names(df) <- df$dateTime
df <- df[,-1]

## remove year (we do not need this)
df <- df[,-4]

str(df)

## Modify R_24hrAvg and R_24hrVariability to int
#df$R_24hrAvg <- as.numeric(df$R_24hrAvg)
#df$R_24hrVariability <- as.numeric(df$R_24hrVariability)
```
# Split inital df into two df's for Ross and Boundary
```{r}
## Split the DF into 1 for Ross and 1 for Boundary
RossDF <- df[,c(1,2,3,4,5,6,7,9,11,13,14,15,16)]
BoundaryDF <- df[,c(1,2,3,4,5,6,8,10,12,13,14,15,17)]

## Let's remove all NAs from these dfs
RossDF <- RossDF[complete.cases(RossDF),]
BoundaryDF <- BoundaryDF[complete.cases(BoundaryDF),]

## Read it to a csv
write.csv(RossDF,"workingRoss.csv")
write.csv(BoundaryDF,"workingBoundary.csv")

## what do these dfs look like?
str(RossDF)
str(BoundaryDF)

## Manually verified there are no NAs/#DIV errors
RossDF <- read.csv("workingRoss_NoNA.csv", header = TRUE)
row.names(RossDF) <- RossDF$X
RossDF <- RossDF[,-1]
#RossDF$R_24hrVariability <- as.numeric(RossDF$R_24hrVariability)
BoundaryDF <- read.csv("workingBoundary.csv", header = TRUE)
row.names(BoundaryDF) <- BoundaryDF$X
BoundaryDF <- BoundaryDF[,-1]

## Let's confirm what these dfs look like...
str(RossDF)
str(BoundaryDF)
```
# Split the Ross and Boundary DF to identify the no, bad, and good class variable
```{r}
## We want to remove out some of the variables from these two DFs. 
## Net_24hrAvg, R/B_24hrVariability, Load_24hrVar, and ross/boundary plant avail...
## These are the variables we will remove to generate our new class variable!

RossClass <- RossDF[,c(6,7,9,12,13)]
RossWorking <- RossDF[,-c(6,9,12,13)]
BoundaryClass <- BoundaryDF[,c(6,7,9,12,13)]
BoundaryWorking <- BoundaryDF[,-c(6,9,12,13)]
str(RossClass)
str(RossWorking)
str(BoundaryClass)
str(BoundaryWorking)
```
## Generate the class level variable
```{r}
## What is the summary of our class-deriving variables?
summary(RossClass)
summary(BoundaryClass)

## If ross/boundaryPlantAvail = 4 or 6, then there is no outage
## If Net, R/B, or Load have 0-25% or 75-100% value, then they are higher volatility - bad outage.

## write to csv and manipulate class in excel (quicker solution)
write.csv(RossClass,"workingRossClass.csv",row.names=FALSE)
write.csv(BoundaryClass,"workingBoundaryClass.csv",row.names=FALSE)

##############################################
## Discretize the quant variables to qualit
#A <- cut(RossClass$Net_24hrAvg, breaks = c(-186791,1457,4299,12944,66286), labels = c("BAD","GOOD","GOOD","BAD"))
#B <- cut(RossClass$R_24hrVariability , breaks = c(-0.3249,-0.0139,-0.0005,0.000343,0.7003), labels = c("BAD","GOOD","GOOD","BAD"))
#C <- cut(RossClass$Load_24hrVar, breaks = c(-0.023,-0.0016,-0.00000331,0.0013,0.0210), labels = c("BAD","GOOD","GOOD","BAD"))
#D <- cut(RossClass$Net_24hrAvg, breaks = c(0,4,5), labels = c("No Outage","Outage"))
#RossClass <- data.frame(A,B,C,D)
#str(RossClass)
##############################################

## was easier to manipulate and assign our class in excel. 
RossLabel <- read.csv("RossLabel.csv",header = TRUE)
RossLabel <- RossLabel[,-2]
BoundaryLabel <- read.csv("BoundaryLabel.csv",header = TRUE)
BoundaryLabel <- BoundaryLabel[,-2]

## Now let's see what these label vectors look like
str(RossLabel)
str(BoundaryLabel)

## Let's drop the label component into our working frames
RossWorking <- data.frame(RossLabel,RossWorking)
BoundaryWorking <- data.frame(BoundaryLabel, BoundaryWorking)

## Remove the rows which have "No Outage". We don't need that!
RossWorking <- RossWorking[!RossWorking$RossLabel == "No Outage",]
BoundaryWorking <- BoundaryWorking[!BoundaryWorking$BoundaryLabel == "No Outage",]

## YIPPEE - WE HAVE OUR DATA FRAMES DONE WITH PREPROCESSING.
str(RossWorking)
str(BoundaryWorking)
```
# Generate the test and train data set
```{r}
## Validation training sets will be a subset of the original.
## Training and validation Ross sets
Revery8rows <- seq(2,nrow(RossWorking),7)
RossTrain <- RossWorking[-Revery8rows,]
RossValidation <- RossWorking[Revery8rows,]
str(RossTrain)
str(RossValidation)

## Training and validation Boundary sets
Bevery8rows <- seq(2,nrow(BoundaryWorking),7)
BoundaryTrain <- BoundaryWorking[-Bevery8rows,]
BoundaryValidation <- BoundaryWorking[Bevery8rows,]
str(BoundaryTrain)
str(BoundaryValidation)
```
# Classifier Techniques
## Discretization / Binary Inputs
```{r}
## Downlaod the libraries we need
library(e1071)
library(naivebayes)
library(data.table)
library(ggplot2)
library(rpart)
library(rattle)
library(rpart.plot)
library(RColorBrewer)
library(Cairo)
library(RWeka)
library(rattle)
library(party)
library(inum)

## We need to transform our data to discretized factors.
## label, hour, month, and day ==> as.factor
BoundaryTrain$BoundaryLabel <- as.factor(BoundaryTrain$BoundaryLabel)
BoundaryTrain$hour <- as.factor(BoundaryTrain$hour)
BoundaryTrain$month <- as.factor(BoundaryTrain$month)
BoundaryTrain$day <- as.factor(BoundaryTrain$day)
BoundaryValidation$BoundaryLabel <- as.factor(BoundaryValidation$BoundaryLabel)
BoundaryValidation$hour <- as.factor(BoundaryValidation$hour)
BoundaryValidation$month <- as.factor(BoundaryValidation$month)
BoundaryValidation$day <- as.factor(BoundaryValidation$day)
RossTrain$RossLabel <- as.factor(RossTrain$RossLabel)
RossTrain$hour <- as.factor(RossTrain$hour)
RossTrain$month <- as.factor(RossTrain$month)
RossTrain$day <- as.factor(RossTrain$day)
RossValidation$RossLabel <- as.factor(RossValidation$RossLabel)
RossValidation$hour <- as.factor(RossValidation$hour)
RossValidation$month <- as.factor(RossValidation$month)
RossValidation$day <- as.factor(RossValidation$day)

## Let's save these as variables in case we run with non discretized
RossTrain_Continuous <- RossTrain
RossValidation_Continuous <- RossValidation
BoundaryTrain_Continuous <- BoundaryTrain
BoundaryValidation_Continuous <- BoundaryValidation

## prchs, sale, flow, flowavg, and load ==> discretized into quartile buckets
## Let's first identify the quartiles we care about and adjust the Boundary dfs
summary(BoundaryTrain[,4:10])

## Variable 1 of 6
BoundaryTrain$MktingPrchs_24hrAvg <- cut(BoundaryTrain$MktingPrchs_24hrAvg, breaks = c(-100000,-7156.4,-6412.0,-2081.9,645),labels=c(1,2,3,4))
BoundaryValidation$MktingPrchs_24hrAvg <- cut(BoundaryValidation$MktingPrchs_24hrAvg, breaks = c(-100000,-7156.4,-6412.0,-2081.9,645),labels=c(1,2,3,4))

## Variable 2 of 6
BoundaryTrain$MktingSale_24hrAvg <- cut(BoundaryTrain$MktingSale_24hrAvg, breaks = c(-1,6827,12490,16968,42536),labels=c(1,2,3,4))
BoundaryValidation$MktingSale_24hrAvg <- cut(BoundaryValidation$MktingSale_24hrAvg, breaks = c(-1,6827,12490,16968,42536),labels=c(1,2,3,4))

## Variable 3 of 6
BoundaryTrain$B_NatFlow <- cut(BoundaryTrain$B_NatFlow, breaks = c(399,11548,16873,20339,97846),labels=c(1,2,3,4))
BoundaryValidation$B_NatFlow <- cut(BoundaryValidation$B_NatFlow, breaks = c(399,11548,16873,20339,97846),labels=c(1,2,3,4))

## Variable 4 of 6
BoundaryTrain$BFlow_24hrAvg <- cut(BoundaryTrain$BFlow_24hrAvg, breaks = c(4170,12401,16851,19892,82699),labels=c(1,2,3,4))
BoundaryValidation$BFlow_24hrAvg <- cut(BoundaryValidation$BFlow_24hrAvg, breaks = c(4170,12401,16851,19892,82699),labels=c(1,2,3,4))

## Variable 5 of 6
BoundaryTrain$Load_Actual <- cut(BoundaryTrain$Load_Actual, breaks = c(662,998,1147,1288,1902),labels=c(1,2,3,4))
BoundaryValidation$Load_Actual <- cut(BoundaryValidation$Load_Actual, breaks = c(662,998,1147,1288,1902),labels=c(1,2,3,4))

## Variable 6 of 6
BoundaryTrain$Load_24hrAvg <- cut(BoundaryTrain$Load_24hrAvg, breaks = c(860,1042.7,1147.1,1242.2,1667.1),labels=c(1,2,3,4))
BoundaryValidation$Load_24hrAvg <- cut(BoundaryValidation$Load_24hrAvg, breaks = c(860,1042.7,1147.1,1242.2,1667.1),labels=c(1,2,3,4))
```
```{r}
## Now, identify the quartiles for the Ross DFs
summary(RossTrain[,4:10])

## Variable 1 of 6
RossTrain$MktingPrchs_24hrAvg <- cut(RossTrain$MktingPrchs_24hrAvg, breaks = c(-189939,-6880.9,-10152.8,-1891.3,909.2),labels=c(1,2,3,4))
RossValidation$MktingPrchs_24hrAvg <- cut(RossValidation$MktingPrchs_24hrAvg, breaks = c(-189939,-6880.9,-10152.8,-1891.3,909.2),labels=c(1,2,3,4))

## Variable 2 of 6
RossTrain$MktingSale_24hrAvg <- cut(RossTrain$MktingSale_24hrAvg, breaks = c(-1,6838,12594,16625,61928),labels=c(1,2,3,4))
RossValidation$MktingSale_24hrAvg <- cut(RossValidation$MktingSale_24hrAvg, breaks = c(-1,6838,12594,16625,61928),labels=c(1,2,3,4))

## Variable 3 of 6
RossTrain$R_NatFlow <- cut(RossTrain$R_NatFlow, breaks = c(350,971,3253,4214,19993),labels=c(1,2,3,4))
RossValidation$R_NatFlow <- cut(RossValidation$R_NatFlow, breaks = c(350,971,3253,4214,19993),labels=c(1,2,3,4))

## Variable 4 of 6
RossTrain$RFlow_24hrAvg <- cut(RossTrain$RFlow_24hrAvg, breaks = c(230,1167.6,3237,3925.4,19565),labels=c(1,2,3,4))
RossValidation$RFlow_24hrAvg <- cut(RossValidation$RFlow_24hrAvg, breaks = c(230,1167.6,3237,3925.4,19565),labels=c(1,2,3,4))

## Variable 5 of 6
RossTrain$Load_Actual <- cut(RossTrain$Load_Actual, breaks = c(664,958,1103,1223,1803),labels=c(1,2,3,4))
RossValidation$Load_Actual <- cut(RossValidation$Load_Actual, breaks = c(664,958,1103,1223,1803),labels=c(1,2,3,4))

## Variable 6 of 6
RossTrain$Load_24hrAvg <- cut(RossTrain$Load_24hrAvg, breaks = c(843,1026,1104,1155.5,1557),labels=c(1,2,3,4))
RossValidation$Load_24hrAvg <- cut(RossValidation$Load_24hrAvg, breaks = c(843,1026,1104,1155.5,1557),labels=c(1,2,3,4))

## Last preprocessing step - remove the labels!
BoundaryTrain_NoLabel <- BoundaryTrain[,-1]
BoundaryTrain_Label <- BoundaryTrain[,1]
BoundaryValidation_NoLabel <- BoundaryValidation[,-1]
BoundaryValidation_Label <- BoundaryValidation[,1]
RossTrain_NoLabel <- RossTrain[,-1]
RossTrain_Label <- RossTrain[,1]
RossValidation_NoLabel <- RossValidation[,-1]
RossValidation_Label <- RossValidation[,1]

## Let's simplify the label column name to be just "label"
colnames(BoundaryTrain)[1] <- "label"
#colnames(BoundaryValidation)[1] <- "label"
colnames(RossTrain)[1] <- "label"
#colnames(RossValidation)[1] <- "label"

## Remove the label from the validation
BoundaryValidation <- BoundaryValidation[,-1]
RossValidation <- RossValidation[,-1]

str(BoundaryTrain)
str(BoundaryValidation)
str(RossTrain)
str(RossValidation)
```
## Decision Tree
```{r}
##################### B O U N D A R Y #########################
## Decision Tree using J48
treeB1 <- J48(label~.,data=BoundaryTrain,control =Weka_control(U=FALSE,M=2,C=0.5))

WOW("J48")

## Use 10 fold cross-validation to evaluate the model
e <- evaluate_Weka_classifier(treeB1,numFolds=10,seed=1,class=TRUE)
e
f <- evaluate_Weka_classifier(treeB1,numFolds=15,seed=500,class=TRUE)
f

## One more tree in J48
treeB2 <- J48(label~.,data=BoundaryTrain,control =Weka_control(U=FALSE,M=5,C=0.5))
g <- evaluate_Weka_classifier(treeB1,numFolds=10,seed=1,class=TRUE)
g

## Apply the model with test dataset
predB<-predict(treeB1,newdata = BoundaryValidation, type=c("class"))
write.csv(predB, file ="C:/Users/ngruet/Desktop/Personal/Grad School/IST 707/project/BoundaryPrediction_DT1.csv")

## Running another tree  with rpart
treeB3 <- rpart(label~., BoundaryTrain)
fancyRpartPlot(treeB3)
predB2 <- predict(treeB3,BoundaryValidation,type="class")
Bresults <- data.frame(predB2,BoundaryValidation_Label)
write.csv(Bresults,"BoundaryPrediction_DT2.csv")
jpeg("treeB3.jpg")

##################### R O S S #########################
## Decision Tree using J48
treeR1 <- J48(label~.,data=RossTrain,control =Weka_control(U=FALSE,M=2,C=0.5))

WOW("J48")

## Use 10 fold cross-validation to evaluate the model
e <- evaluate_Weka_classifier(treeR1,numFolds=10,seed=1,class=TRUE)
e
f <- evaluate_Weka_classifier(treeR1,numFolds=15,seed=500,class=TRUE)
f

## One more tree in J48
treeR2 <- J48(label~.,data=RossTrain,control =Weka_control(U=FALSE,M=5,C=0.5))
g <- evaluate_Weka_classifier(treeR1,numFolds=10,seed=1,class=TRUE)
g

## Apply the model with test dataset
predR<-predict(treeR1,newdata = RossValidation, type=c("class"))
write.csv(predR, file ="C:/Users/ngruet/Desktop/Personal/Grad School/IST 707/project/RossPrediction_DT1.csv")

## Running another tree  with rpart
treeR3 <- rpart(label~., RossTrain)
fancyRpartPlot(treeR3)
predR2 <- predict(treeR3,RossValidation,type="class")
Rresults <- data.frame(predR2,RossValidation_Label)
write.csv(Rresults,"RossPrediction_DT2.csv")
jpeg("treeR3.jpg")
```
## NaiveBayes
```{r}
## Let's check our train, validation, and test datasets
str(BoundaryTrain)
str(BoundaryValidation)
str(RossTrain)
str(RossValidation)

################ B O U N D A R Y ####################
## Run the e1071 NB!
NB_Btrain <- naiveBayes(label~., data=BoundaryTrain,na.action=na.pass)
NB_Btrain_pred <- predict(NB_Btrain,BoundaryValidation)

table(NB_Btrain_pred,BoundaryValidation_Label)

## Rate of misclassification error
(468+430)/(2313+468+430+4517)

###################### R O S S #######################
## Run the e1071 NB!
NB_Rtrain <- naiveBayes(label~., data=RossTrain,na.action=na.pass)
NB_Rtrain_pred <- predict(NB_Rtrain,RossValidation)

table(NB_Rtrain_pred,RossValidation_Label)

## Rate of misclassification error
(112+129)/(1069+3095+112+129)

```